Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States.

### 2025-04-11 code/analysis-code/statistical-analysis-v2.qmd
Part 4 additions

Fit the Interaction Model
```{r}
# GLM with 3-way interaction to explore mortality trends by Year * race_ethnicity * sex, stratified by age group

## ---- load_glm_race_sex_age_model, message=TRUE, warning=FALSE -------------------

glm_path <- here::here("results", "output", "glm_race_sex_age_model.rds")

if (file.exists(glm_path)) {
  message("Loading saved GLM model with 3-way interaction...")
  glm_race_sex_age <- readRDS(glm_path)
} else {
  message("Fitting GLM model with 3-way interaction...")
  glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group,
                          data = county_race_sex, family = gaussian())
  saveRDS(glm_race_sex_age, glm_path)
}

#glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group, data = processed_data_no_percent, family = gaussian())

# Save model
#saveRDS(glm_race_sex_age, here::here("results", "output", "glm_race_sex_age_model.rds"))

@@

```{r}
# Define file paths for saving/loading tuning results
tuned_path     <- here::here("results", "large-files", "rf_tuned.rds")
metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
best_param_path <- here::here("results", "output", "rf_best_params.rds")

@@

  message("Tuning complete and results saved.")}

```
```{r}
## ---- fit_or_load_rf_model, message=TRUE, warning=FALSE -------------------

use_slim_model <- FALSE # Toggle this to FALSE to regenerate slim model if needed

slim_rf_path <- here::here("results", "output", "rf_model_minimal.rds")
full_rf_path <- here::here("results", "large-files", "rf_tuned.rds")
full_rf_metrics_path <- here::here("results", "output", "rf_tuned_metrics.rds")
best_rf_params_path <- here::here("results", "output", "rf_best_params.rds")

if (use_slim_model) {
  if (!file.exists(slim_rf_path)) {
    stop("Slimmed model not found. Please run with use_slim_model <- FALSE first to generate it.")
  }

  message("Loading slimmed RF model...")
  rf_slim <- readRDS(slim_rf_path)
  rf_final_model <- rf_slim$model
  rf_metrics <- rf_slim$metrics
  rf_vip <- rf_slim$vip
  best_rf_params <- readRDS(best_rf_params_path)

} else {
  if (!file.exists(full_rf_path)) {
    stop("Full Random Forest model not found. Please rerun the tuning block.")
  }

  message("Loading full RF model and generating slim version...")
  rf_tuned <- readRDS(full_rf_path)
  rf_metrics <- readRDS(full_rf_metrics_path)
  rf_final_model <- extract_workflow(rf_tuned)
  rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()
  best_rf_params <- readRDS(best_rf_params_path)

  if (!file.exists(slim_rf_path)) {
    saveRDS(
      list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
      file = slim_rf_path
    )
    message("Slimmed model saved.")}}

```
Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.

```{r}
#Slim version already loads rf_metrics, so no need to reload
rf_metrics %>% filter(.metric == "rmse") %>% arrange(mean)

@@

# Plot RMSE across combinations
rf_metrics %>% filter(.metric == "rmse") %>%
  ggplot(aes(x = factor(mtry), y = mean, color = factor(min_n))) +
  geom_point(size = 3) +
  geom_line(aes(group = min_n)) +
  labs(title = "Random Forest Tuning Results (RMSE)",
       x = "mtry (Number of Predictors)",
       y = "Mean RMSE",
       color = "min_n") +
  theme_minimal()

@@

  # Extract best parameters
  best_params <- best_rf_params

@@

# File paths
lasso_tuned_path <- here::here("results", "large-files", "lasso_tuned.rds")
lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
lasso_best_param_path <- here::here("results", "output", "lasso_best_params.rds")

# Load or tune
if (file.exists(lasso_tuned_path) && file.exists(lasso_metrics_path) && file.exists(lasso_best_param_path)) {
  message("Loading saved LASSO tuning results...")

  lasso_tuned <- readRDS(lasso_tuned_path)
  lasso_tuned_metrics <- readRDS(lasso_metrics_path)
  best_lasso_params <- readRDS(lasso_best_param_path)
} else {
  message("Running LASSO tuning...")

@@

  set.seed(123)
  lasso_tuned <- tune_grid(
    lasso_workflow,
    resamples = cv_folds,
    grid = lasso_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE))

@@

  # Save results
  saveRDS(lasso_tuned, lasso_tuned_path)
  saveRDS(lasso_tuned_metrics, lasso_metrics_path)
  saveRDS(best_lasso_params, lasso_best_param_path)

  message("LASSO Tuning complete and results saved.")
}

# Preview metrics
print(lasso_metrics)
```

Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
```{r}
## ---- slim_lasso_model_save, message=TRUE, warning=FALSE -------------------

slim_lasso_path <- here::here("results", "output", "lasso_model_minimal.rds")

if (!file.exists(slim_lasso_path)) {
  message("Creating slimmed LASSO model object...")

  lasso_final <- extract_workflow(lasso_tuned)
  lasso_metrics <- lasso_tuned_metrics

  # Extract coefficients
  lasso_fit <- extract_fit_parsnip(lasso_final)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  saveRDS(
    list(model = lasso_final, metrics = lasso_metrics, coefficients = lasso_coeffs),
    file = slim_lasso_path
  )

  message("Saved slimmed LASSO model as lasso_model_minimal.rds")
} else {
  message("Slimmed LASSO model already exists. Skipping save.")
}

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.
```{r}
## ---- load_lasso_model, message=TRUE, warning=FALSE -------------------

use_slim_lasso_model <- TRUE  # Toggle to FALSE to use full model

if (use_slim_lasso_model) {
  if (!file.exists(here::here("results", "output", "lasso_model_minimal.rds"))) {
    stop("Slim LASSO model not found. Please run the slim-save chunk or disable slim mode.")
  }

  message("Loading slimmed LASSO model...")
  lasso_slim <- readRDS(here::here("results", "output", "lasso_model_minimal.rds"))

  lasso_final_model <- lasso_slim$model
  lasso_metrics <- lasso_slim$metrics
  lasso_coeffs <- lasso_slim$coefficients
  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))

} else {
  if (!file.exists(here::here("results", "large-files", "lasso_tuned.rds"))) {
    stop("Full LASSO model not found.")
  }

  message("Loading full LASSO model...")
  lasso_tuned <- readRDS(here::here("results", "large-files", "lasso_tuned.rds"))
  lasso_metrics <- readRDS(here::here("results", "output", "lasso_tuned_metrics.rds"))
  lasso_final_model <- extract_workflow(lasso_tuned)

  lasso_fit <- extract_fit_parsnip(lasso_final_model)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))}

```{r}
#Plot LASSO Tuning Plot
# Plot RMSE vs penalty (log10 scale)
lasso_tune_plot <- lasso_metrics %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = penalty, y = mean)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line() +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "LASSO Tuning Results (RMSE vs Penalty)",
       x = "Penalty (log10 scale)",
       y = "Mean RMSE")

>>>>

Fit the Interaction Model
```{r}

# GLM with 3-way interaction to explore mortality trends by Year * race_ethnicity * sex, stratified by age group
glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group,
                        data = processed_data_no_percent,
                        family = gaussian())

# Save model
saveRDS(glm_race_sex_age, here::here("results", "output", "glm_race_sex_age_model.rds"))

@@

```{r}
# Define file paths for saving/loading tuning results
tuned_path     <- here::here("results", "output", "rf_tuned.rds")
metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
best_param_path <- here::here("results", "output", "rf_best_params.rds")

@@

  message("Tuning complete and results saved.")
}

# Preview metrics
print(rf_tuned_metrics)

# Load metrics (if not already loaded)
rf_tuned_metrics <- readRDS(here::here("results", "output", "rf_tuned_metrics.rds"))

# Sort by RMSE to find best performing combo
rf_tuned_metrics %>% filter(.metric == "rmse") %>% arrange(mean)

@@

# Plot RMSE across combinations
rf_tuned_metrics %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = factor(mtry), y = mean, color = factor(min_n))) +
  geom_point(size = 3) +
  geom_line(aes(group = min_n)) +
  labs(title = "Random Forest Tuning Results (RMSE)",
       x = "mtry (Number of Predictors)",
       y = "Mean RMSE",
       color = "min_n") +
  theme_minimal()

@@

  # Extract best parameters
  best_params <- select_best(rf_tuned, metric = "rmse")

@@

# File paths
lasso_tuned_path <- here::here("results", "output", "lasso_tuned.rds")
lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
lasso_best_path <- here::here("results", "output", "lasso_best_params.rds")

# Load or tune
if (file.exists(lasso_tuned_path) &&
    file.exists(lasso_metrics_path) &&
    file.exists(lasso_best_path)) {
  message("Loading saved LASSO tuning results...")
  lasso_tuned <- readRDS(lasso_tuned_path)
  lasso_tuned_metrics <- readRDS(lasso_metrics_path)
  best_lasso_params <- readRDS(lasso_best_path)
} else {
  message("Running LASSO tuning...")

@@

  set.seed(123)
  lasso_tuned <- tune_grid(
    lasso_workflow,
    resamples = cv_folds,
    grid = lasso_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE)
  )

@@

  # Save results
  saveRDS(lasso_tuned, lasso_tuned_path)
  saveRDS(lasso_tuned_metrics, lasso_metrics_path)
  saveRDS(best_lasso_params, lasso_best_path)

  message("Tuning complete and results saved.")
}

# Preview metrics
print(lasso_tuned_metrics)
```

```{r}
#Plot LASSO Tuning Plot
# Plot RMSE vs penalty (log10 scale)
lasso_tune_plot <- lasso_tuned_metrics %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = penalty, y = mean)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line() +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "LASSO Tuning Results (RMSE vs Penalty)",
       x = "Penalty (log10 scale)",
       y = "Mean RMSE")

### 2025-04-11 code/processing-code/processingfile-v1.qmd
Part 4 additions

---
title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA"
author: "Shaun van den Hurk"
date: "2025-02-20"
output: html_document
---

@@

Stratify by country-level overall:
```{r}
#Using the filter function to help us with stratification
county_overall <- hypertens_adult_filtered_no_percent %>%
  filter(sex == "Overall", race_ethnicity == "Overall") %>% # Keep only total county-level data
  select(LocationID, LocationDesc, Year, mortalities, X_long, Y_lat, LocationAbbr, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group)

@@

Running some EDA and checking summaries of data:

We are going to make use of loops and creating a list of the different datasets that we created to reduce separate code to run the descriptive statisitics for each dataset. We are not include the percent change dataset in these evaluations.

Creating this datset list:
```{r}
list_datasets <- list(
  "Full Dataset" = hypertens_adult_filtered_no_percent,
  "County Race-Sex" = county_race_sex,
  "County Race" = county_race,
  "County Sex" = county_sex,
  "County Overall" = county_overall
)

We will run the descriptive statistics for each dataset (as a single loop)
```{r}
# Loop through each dataset and print basic statistics
summary_stats <- bind_rows(lapply(names(list_datasets), function(name) {
  stats <- list_datasets[[name]] %>%
    summarise(
      mean = mean(mortalities, na.rm = TRUE),
      median = median(mortalities, na.rm = TRUE),
      sd = sd(mortalities, na.rm = TRUE),
      min = min(mortalities, na.rm = TRUE),
      max = max(mortalities, na.rm = TRUE)
    )
  stats$Dataset <- name  # Add dataset name for identification
  return(stats)
}))

# Print summary statistics
tibble::as_tibble(summary_stats) %>% 
  relocate(Dataset) %>% 
  print()
```

# Identify potential missing values in each dataset
```{r}
cat("\n### Missing Data Summary:\n")
lapply(names(list_datasets), function(name) {
  missing_count <- sum(is.na(list_datasets[[name]]))
  cat(name, "has", missing_count, "missing values.\n")
})
```

We want to visualise the different datasets:

We will revisit the histogram with the percent change removed.
```{r}
# Visualize the distribution of mortality rates with a histogram
revised_mortalities_histogram <- ggplot(hypertens_adult_filtered_no_percent, aes(x = mortalities)) +
  geom_histogram(binwidth = 20, fill = "blue", alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, max(hypertens_adult_filtered_no_percent$mortalities, na.rm = TRUE), by = 200)) +
  theme_minimal() +
  labs(title = "Distribution of Mortality Rates", x = "Mortality Rates", y = "Count")

saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
# Print the mortalities histogram
print(revised_mortalities_histogram)
```

This is the total dataset and so the data is similar to the previous histogram. There is a strong right skew and long tail. Peak mortalities are below 100, and the total values likely account for the second peak.

We will use a boxlot to get an overview of the other data:

```{r}
# Compare mortality rates across race-sex and sex stratifications

boxplot_race_sex <- ggplot(county_race_sex, aes(x = race_ethnicity, y = mortalities, fill = sex)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Race and Sex", x = "Race/Ethnicity", y = "Mortality Rate")

saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))

boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")

saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))

The box plots provide a good overview of the data nd indicate that there seem to be distinct trends and differences between races. It also shows that there are differences between sexes but these are less marked and distinct than between races.
We will ook at this in the statistical analysis.

We want an idea on any sptial trends in the data. We are using the sf function to plot the mortality data by the latitude and longitude.
```{r}
# Convert data to spatial format and create a map of mortality rates
# Ensure longitude and latitude values are numeric
hypertens_adult_filtered_no_percent <- hypertens_adult_filtered_no_percent %>%
  mutate(X_long = as.numeric(X_long), Y_lat = as.numeric(Y_lat))
# Convert to spatial object
sf_data <- st_as_sf(hypertens_adult_filtered_no_percent, coords = c("X_long", "Y_lat"), crs = 4326)

# Generate a spatial plot of mortality rates
ggplot(sf_data) +
  geom_sf(aes(color = mortalities)) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Geospatial Distribution of Mortality Rates")

This is a very nice overview and can help us to evauate the trends further or to work on a more intricate plot. We must remember that we have not separated groups or totals yet.

We want an overview of any temporal (time based) trends in the data. We will provide a proof of concept using a single county due to the overlap of data that will occur otherwise. The statistical analysis will include further refinements. We are only working with the values where sex and race were both overall to reduce overlap.
```{r}
# Line plot of mortality rates over time
# Select a single county to illustrate the trend
selected_county <- "13037"  # County Location ID for Calhoun, GA
county_subset <- county_overall %>% filter(LocationID == selected_county)

temporal_trends_single_county <- ggplot(county_subset, aes(x = Year, y = mortalities, group = LocationID)) +
  geom_line(alpha = 0.6, color = "blue") +
  geom_smooth(se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = paste("Mortality Rate Trends in Calhoun"), 
       x = "Year", y = "Mortality Rate")

print(temporal_trends_single_county)
```

## Key Findings and Next Steps

- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
- Mortality rates vary markedly by race and sex, justifying stratified analysis.
- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
- Temporal trend indicate a general increase the selected county, which should be modeled.
- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.

>>>>

---
title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA - Processing script"
author: "Shaun van den Hurk"
date: "2025-02-20"
output: html_document
---

@@

Stratify by country-level overall:
```{r}
#Using the filter function to help us with stratification
county_overall <- hypertens_adult_filtered_no_percent %>%
  filter(sex == "Overall", race_ethnicity == "Overall") %>% # Keep only total county-level data
  select(LocationID,Year,LocationAbbr, LocationDesc,  mortalities, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group, X_long, Y_lat)

@@

## Summary and Next Steps

- The data was cleaned and processed to result in a dataset and a few different variables that are much smaller and more manageable to work with.
- We checked for missing variables and checked overall trends and removed d data and variables that were not useful.
- We have saved the different created variables as a list saved into one variable.
- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in exploratory data analysis.

### 2025-04-11 products/manuscript/Manuscript.qmd
Part 4 additions

---
title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
author: "Shaun van den Hurk"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

* First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
* Second Author$^{3, *}$
* Third Autor$^{3, \dagger}$
* Last Author$^{1, 2, \land}$

@@

$\land$ Corresponding author: some@email.com

@@

# Summary/Abstract
Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States. This project focuses on hypertension-related cardiovascular disease and hopes to help provide a greater understandign of the dynamics and patterns that may exist with hypertension and cardiovascualr diseased related mortality in the United States.
The project seeks to evaluate data from the CDC on the rates of hypertension-related cardiovascular disease mortality and the associated demographic factors for this collected data, including county, age group, race/ethnicity, and sex. The data represents information collected from 2000 to 2019. This will be used to see if there are any potential patterns or correlation between the rate of mortality and these other demographic factors within the data.

@@

@@

@@

A summary table was produced using the skimr function which is printed below.

```{r}
skim_summary_hypertens_filtered <- readRDS(here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
# Print the skim summary
print(skim_summary_hypertens_filtered)

```

Our summary data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.

A histogram was made of the mortalities columns since this is the main (only) column of interest with numerical data.

```{r}
revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
# Print the mortalities histogram
plot(revised_mortalities_histogram)
```

Our histogram depicts the hypertension-related cardiovasculat disease mortalities in the dataset, showing the number of mortalities and the frequency (count) of sprecific mortality values.
The mortality counts are clustered towards the lower end of the scale, with a strong right-skew (long tail to the right/higher values). Peak mortalities are below 100, and the total values likely account for the second peak.

Boxplots were created to evaluate the mortalities by race and by sex.

@@

## Key Findings and Next Steps

- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
- Mortality rates vary markedly by race and sex, justifying stratified analysis.
- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
- Temporal trend indicate a general increase the selected county, which should be modeled.
- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.

@@

```{r}
#Print the glm plot
glm_interaction_plot <- image_read(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"))
print(glm_interaction_plot)
```

```{r}
#Print the combined spatial plot
combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
print(combined_spatial_plot)
```

@@

_

@@

>>>>

---
title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
author: "Shaun van den Hurk"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
editor: 
  markdown: 
    wrap: 72
---

@@

-   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Autor$^{3, \dagger}$
-   Last Author$^{1, 2, \land}$

@@

$\land$ Corresponding author: some\@email.com

@@

# Summary/Abstract

Lifestyle diseases such as cardiovascular disease make up the highest
causes of death in developed nations such as the United States. This
project focuses on hypertension-related cardiovascular disease and hopes
to help provide a greater understandign of the dynamics and patterns
that may exist with hypertension and cardiovascualr diseased related
mortality in the United States. The project seeks to evaluate data from
the CDC on the rates of hypertension-related cardiovascular disease
mortality and the associated demographic factors for this collected
data, including county, age group, race/ethnicity, and sex. The data
represents information collected from 2000 to 2019. This will be used to
see if there are any potential patterns or correlation between the rate
of mortality and these other demographic factors within the data.

@@

Study Aims:
- Aim 1: Assess national temporal trends in hypertension-related CVD mortality, stratified by age group.
- Aim 2: Identify spatial clustering of county-level mortality.
- Aim 3: Evaluate disparities in mortality trends by race/ethnicity and sex, stratified by age group.
- Aim 4: Build predictive models using demographic and geographic features to estimate mortality trends.

Hypotheses:
- Mortality trends vary significantly over time and across age groups.
- County-level mortality exhibits spatial clustering, reflecting geographic disparities.
- Temporal trends differ significantly across race and sex, with widening inequities over time.
- Predictive models can identify key demographic and spatial factors associated with higher mortality trends.

@@

Aim 1: We assessed national temporal trends in hypertension-related
cardiovascular disease (CVD) mortality from 2000 to 2019 using a series
of regression models. First, a linear regression model was fit to assess
the overall relationship between mortality rates and year. Next,
generalized linear models (GLMs) were used to evaluate time trends
stratified by race/ethnicity, sex, and age group. Interaction terms
(e.g., Year × race/ethnicity) were included to test for differential
trends across subgroups. Estimated annual changes in mortality rates
were compared across age groups using the emtrends and contrast
functions from the emmeans package.

Spatial Patterns of Mortality Aim 2: Spatial clustering analysis:
County-level mortality data were converted to spatial features using
geographic coordinates. Global and Local Moran’s I statistics were used
to assess spatial autocorrelation. Spatial weights were generated using
5-nearest-neighbor methods. Local clustering was visualized using
standardized mortality z-scores and LISA cluster classifications.

Evaluating National Demographic Disparities Aim 3: We fit generalized
linear models (GLMs) with interaction terms to estimate yearly changes
in hypertension-related CVD mortality rates from 2000–2019 across
demographic groups. Using emtrends() from the emmeans package, we
extracted slope estimates by race/ethnicity and sex, stratified by age
group (35–64 and 65+). Pairwise contrasts were conducted to evaluate
statistically significant differences in trends between groups.

Statistical Modeling (Aim 4) To predict county-level
hypertension-related cardiovascular disease (CVD) mortality, we trained
two supervised machine learning models: a random forest and a LASSO
regression. Models were trained on 80% of the data, with the remaining
20% held out for testing. Ten-fold cross-validation was used within the
training set for hyperparameter tuning. Predictors included year, age
group, race/ethnicity, and sex. The modeling dataset included demographic predictors (age group, race/ethnicity, sex) and time (year). Since several predictors were categorical, we applied dummy encoding (one-hot encoding) using step_dummy() within the recipes framework to convert these factors into binary indicator variables. We also normalized numeric predictors and removed any zero-variance predictors. Models were trained using a stratified 80/20 train/test split and evaluated using 10-fold cross-validation on the training set.
Random forest hyperparameters were selected from a regular grid of mtry and
min_n values, while LASSO regression used a log-scaled penalty grid.
Final model performance was evaluated using RMSE, MAE, and R² on the
test set. Confidence intervals were not computed for these metrics to
avoid additional computational burden.

@@

We began by exploring patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S. adults aged 35 and older from 2000 to 2019. The EDA included summary statistics, histograms, and boxplots to assess mortality distributions overall and by key demographic subgroups (race/ethnicity, sex, age group). Temporal trends were visualized using line plots stratified by subgroup and geographic region. Spatial variation was explored with choropleth maps displaying county-level mortality rates across time and demographic strata. These descriptive analyses revealed substantial variation in mortality trends over time, with emerging disparities by race, sex, and age group. Additionally, spatial visualizations suggested clustering in certain states, laying the foundation for formal spatial and demographic analyses in later aims. All visualizations and cleaned summary tables were saved for downstream modeling.

@@

@@

```{r, echo=FALSE, message=FALSE}
#Print the glm plot
glm_interaction_plot <- image_read(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"))
print(glm_interaction_plot)
```

```{r, echo=FALSE, message=FALSE}
#Print the combined spatial plot
combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
print(combined_spatial_plot)
```

@@

Temporal Trends in Mortality Rates (2000–2019) (Aim 1): From 2000 to
2019, overall CVD mortality rates showed minimal change at the national
level. However, stratified models revealed substantial differences
across subgroups. Black individuals consistently exhibited the highest
mortality rates, though a declining trend was observed. In contrast,
rates among American Indian/Alaska Native individuals increased markedly
over time. Older adults (65+) had significantly higher mortality rates
and experienced a steeper upward trend compared to adults aged 35–64
(estimated slope: 1.22 vs. 1.04 deaths per 100,000 per year; p \<
0.001). These findings highlight persistent disparities and suggest
diverging subgroup trends in hypertension-related CVD mortality.

Spatial clustering analysis (Aim 2): We evaluated spatial patterns in
age-adjusted county-level hypertension-related cardiovascular disease
(CVD) mortality using global and local spatial autocorrelation methods.
Global spatial autocorrelation of county-level mortality rates was
statistically significant (Moran’s I = 0.0435, p \< 0.001), indicating a
weak but non-random spatial clustering pattern across the United States.
Local Indicators of Spatial Association (LISA) identified significant
clusters of high and low mortality values. LISA cluster types (e.g.,
high-high, low-low, spatial outliers) were visualized using
centroid-based maps for the contiguous U.S. and select states (Georgia,
Mississippi, Texas, Kentucky). Despite the overall geographic patterns,
only a small number of statistically significant local clusters were
detected, suggesting that while some regions exhibit elevated risk,
strong clustering at the county level may be limited or diffuse.

Investigating demographic disparities (Aim 3): Annual changes in
mortality rates varied substantially across demographic groups. Among
adults aged 35–64, Black individuals experienced the steepest declines,
while mortality increased most among males and American Indian/Alaska
Native populations. Similar patterns were observed among those aged 65+,
with significant disparities between groups. Pairwise comparisons
revealed the largest slope differences between Overall Male and Black
Overall groups, followed by disparities involving American Indian/Alaska
Native populations. These findings suggest widening inequities in
hypertension-related CVD mortality trends.

Predictive Modeling of County-Level Mortality (Aim 4) To evaluate the
predictive capacity of demographic and temporal factors on
hypertension-related cardiovascular mortality, we compared two machine
learning models: a random forest and a LASSO regression. Both models
were trained on 80% of the dataset and evaluated on a held-out 20% test
set, using cross-validation during tuning to optimize performance.

The random forest model demonstrated the strongest performance,
achieving a Root Mean Squared Error (RMSE) of 115.3, a Mean Absolute
Error (MAE) of 66.9, and an R² of 0.748 on the test set. In comparison,
the LASSO regression model yielded an RMSE of 119.6, MAE of 68.1, and R²
of 0.726. While slightly less accurate, the LASSO model offers valuable
interpretability by shrinking less important coefficients to zero,
effectively performing variable selection.

When comparing predicted vs observed plots, both models exhibited a distinct pattern of horizontal banding. This is likely due to the large number of categorical predictors (race, sex, age group) which were dummy-encoded. Many subsets of the population across counties share identical predictor combinations (e.g., "Black females aged 65+ in 2002"), leading to repeated predicted values for those groups. This results in stacked horizontal lines in the prediction plots, especially when using models like LASSO that produce linear fits with limited flexibility, and even in random forests when groupings dominate feature structure.

A comparison of model performance is summarized in Table (...X...).
These results suggest that both models capture meaningful patterns in
the data, but the random forest model more accurately predicts mortality
outcomes. Importantly, the use of interpretable models like LASSO can
aid in identifying potential risk factors despite modest reductions in
predictive power.

```{r, echo=FALSE, message=FALSE}
#Print the model performance summary
model_performance_summary_table <- readRDS(here::here("results", "tables", "model_performance_summary.rds"))
model_performance_summary_table
```

\_

@@

Aim 3: These results show that hypertension-related CVD mortality trends
have not improved equally across demographic groups. While some
populations saw modest or no change, others—particularly Black
adults—experienced significant declines. Conversely, mortality rates
rose among males and American Indian/Alaska Native individuals. The
magnitude and significance of these differences reinforce the need for
targeted public health efforts to reduce disparities in cardiovascular
outcomes.
